Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-03-28 paper/manuscript.Rmd
Begins to take final Colin's comments

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(d) Sorbonne Université, CNRS, IRD, INRAE, Université Paris Est Créteil, 15
Institute of Ecology and Environmental Sciences of Paris (iEES-Paris), Paris,
France

@@

- Keywords: food web structure, biodiversity changes, redundancy, biodiversity &
  ecosystem functioning

# Abstract (229 / 250 words max)

Biodiversity temporal changes are expected to profoundly impact ecosystem
functioning and stability, yet large-scale empirical evidence remains limited. Bridging this
gap requires aligning biodiversity temporal trends with food web theory and
Biodiversity-Ecosystem Functioning research, which provide strong predictions
about the ecological consequences of biodiversity loss. Most insights rely on
spatial comparisons of ecological communities, assuming they reflect temporal
trends—a space-for-time substitution. Here, we analyze biodiversity and food web
time series from over 400 riverine fish communities across France (1995–2018) to
evaluate how biodiversity temporal trends reshape food web structure. We examine
relationships among species richness, community biomass, and three key food web
metrics: connectance, weighted average trophic level, and trophic pathway
redundancy. We also test the space-for-time hypothesis using spatial gradients
and a theoretical food web model. Structural equation models and robustness
analyses reveal that declines in species richness strongly correlate with
reduced community biomass, with top trophic-level species playing a
disproportionate role in food web changes. Declining biomass is associated with the loss
of top trophic levels and decreased connectance, while species richness
negatively correlates with connectance. Both biomass and species richness
declines reduce trophic pathway redundancy, suggesting biodiversity loss
undermines community robustness to future perturbations. Our findings align with
food web theory and Biodiversity-Ecosystem Functioning research, support the
validity of space-for-time approaches for basic food web metrics, and highlight
the utility of food web metrics as indicators of ecosystem function and
fragility.

# Significance (118 / 120 words max)

Understanding how biodiversity loss affects ecological communities is crucial
for predicting and mitigating human impacts on ecosystems. This study provides
comprehensive evidence on how biodiversity changes are reshaping food web structure using
an extensive dataset of riverine fish communities. We demonstrate that changes
in species richness and community biomass consistently alter trophic levels,
connectance, and the redundancy of feeding relationships. Our findings show how
energetic constraints and biodiversity-ecosystem functioning mechanisms jointly
shape food web assembly and disassembly. Importantly, we show that declining
biodiversity reduces the redundancy of trophic pathways, potentially making
communities more vulnerable to future perturbations. These results suggest that
food web structure metrics could serve as valuable indicators for monitoring
ecosystem health and evaluating conservation strategies.

# Introduction (487 words)

Human pressures are driving profound biodiversity changes, including freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024] and reductions in
marine fish biomass due to intensified fishing pressures on predatory species
[@christensen_century_2014]. A central goal of Biodiversity-Ecosystem
Functioning and food web research has been to understand the ecological
consequences of these widespread biodiversity losses for ecosystem functioning
and stability [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @schneider_animal_2016; @dunne_network_2002;
@sole_complexity_2001; @allesina_functional_2009; @montoya_press_2009; @sanders_trophic_2018]. However,
research on biodiversity changes remains largely disconnected from studies
examining their ecological impacts, primarily due to a lack of simultaneous
temporal data on biodiversity and food web dynamics [@danet_species_2021]. As a
result, much of our understanding of the ecological consequences of biodiversity
changes over time relies on experimental and theoretical approaches. These
approaches often rely on the critical assumption that spatial comparisons of
ecological communities can serve as a proxy for temporal trends—a space-for-time
substitution—when direct temporal data are scarce. To advance our understanding
of the ecological consequences of biodiversity changes, it is essential to
investigate how these changes are reshaping food web structure in natural
settings.

Here, we bridge the gap between biodiversity changes and food web dynamics by
leveraging a unique dataset of riverine fish communities sampled repeatedly over
a 23-year period (1995–2018) across 403 locations in France. By reconstructing
yearly food web structures [@danet_species_2021; @bonnaffe_comparison_2021], we
investigate how temporal biodiversity trends reshape food web metrics such as
connectance, weighted average trophic level, and the proportion of redundant
trophic links (hereafter “redundancy”). Linking these findings to established
theories of Biodiversity-Ecosystem Functioning and food web dynamics, we address
three key objectives: (i) documenting how temporal biodiversity trends reshape
food web structure, (ii) connecting these changes to ecosystem functioning and
food web theories, and (iii) assessing the validity of the space-for-time
substitution assumption in the context of biodiversity and food web structure
changes.

The structure of ecological networks, which describes the number, distribution,
and strength of interactions among species, is critical for understanding
species coexistence, ecosystem functioning, and stability to perturbations
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024]. Both
theory and empirical evidence demonstrate that species richness and food web
structure influence coexistence [@chesson_interaction_2008], ecosystem
productivity [@thebault_food-web_2003; @wang_biodiversity_2018], temporal
stability [@danet_species_2021; @danet_response_2024;
@eschenbrenner_diversity_2023], and robustness to disturbances
[@sole_complexity_2001; @dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @montoya_press_2009; @sanders_trophic_2018]. For instance,
keystone predators can regulate dominant competitors, preventing competitive
exclusion and maintaining species richness [@heithaus_predicting_2008;
@johnson_rarity_2007; @estes_trophic_2011]. Similarly, food webs with higher
trophic levels are associated with greater ecosystem productivity
[@wang_biodiversity_2018; @schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023] and community stability
[@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024],
while the number and distribution of trophic links determine resilience to
perturbations [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009;
@allesina_functional_2009; @sanders_trophic_2018]. Understanding how
biodiversity changes reshape ecological networks is therefore essential for
quantifying and mitigating anthropogenic impacts on ecosystems.

Three key components link biodiversity changes to food web structure. First, the
energetic paradigm of food web dynamics posits that the presence of top
consumers depends on sufficient resource availability at lower trophic levels,
driven by bottom-up productivity [@lindeman_trophic-dynamic_1942;
@gravel_trophic_2011]. Thus, the biomass of generalist top consumers is likely
constrained by overall community biomass. Second, connectance—the proportion of
realized trophic links—typically decreases with increasing species richness, as
consumers are limited in the number of resources they can exploit
[@oconnor_general_2017; @dunne_network_2006]. However, higher species richness
is expected to increase the redundancy of trophic pathways, enhancing ecosystem
robustness to species extinctions [@dunne_network_2002; @allesina_who_2004;
@bodini_using_2009; @allesina_functional_2009]. This redundancy underpins the
insurance effect of biodiversity, where greater diversity buffers ecosystems
against perturbations and environmental changes [@yachi_biodiversity_1999;
@loreau_ecological_2022]. Finally, biodiversity-ecosystem functioning theory
suggests that species richness promotes community biomass through niche
complementarity [@loreau_biodiversity_2001]. In food webs, the presence of
higher trophic levels reduces niche overlap among producers, further boosting
productivity and biomass [@schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023]. Together, these mechanisms highlight how
biodiversity changes can alter trophic link distribution, robustness, and
ecosystem functioning.

To achieve these objectives, we analyzed temporal trends in species richness,
biomass, and food web structure across `r nb_op` community samplings (Fig.
\@ref(fig:matmet)a-c, Table S1). Using structural equation modeling (SEM), we
inferred the relative effects of biomass and species richness trends on food web
metrics, as well as the effect of species richness on biomass (Fig.
\@ref(fig:matmet)d). We also conducted a robustness analysis on the meta-food
web (comprising all trophic species and their interactions) to identify species
extinction scenarios that could explain observed variations in food web
structure. To test the space-for-time hypothesis, we compared temporal findings
with spatial gradients using two approaches: (i) replicating our temporal
analysis using median biodiversity and food web metrics across sites (Fig.
\@ref(fig:matmet)c & e), and (ii) simulating food web dynamics across gradients
of species richness and primary productivity using a bioenergetic model
[@brose_allometric_2006; @lajaaiti_ecologicalnetworksdynamicsjl_nodate]. These
simulations allowed us to evaluate whether theoretical predictions align with
our empirical data, which are limited to fish biodiversity.

```{r}
fig_mat_cap <- paste0("Methodological Approach: (a) We analyzed ", nb_st, "
  stream fish communities in France, each monitored for at least 10 years. (b)
  For each site and sampling event, we reconstructed food webs using a metaweb
  based on ontogenetic diet information and predator-prey body size ratios. (c)
  We computed temporal trends and median values for community biomass, species
  richness, connectance, average trophic level, and redundancy at each site. (d)
  We then assessed relationships among these food web descriptors over time and
  (e) across spatial gradients.")
```

```{r matmet, fig.cap=fig_mat_cap, out.width="100%"}
knitr::include_graphics(get_mypath("figures", "mat_met.png"))
```

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species
richness, biomass, connectance, weighted average trophic level, and the
proportion of redundant trophic links (hereafter “redundancy”, see Appendix B)
from our `r nb_st` fish communities with 10+ years of population dynamics
spanning 1995-2018 over France (Fig. \@ref(fig:matmet)a-c). We implemented
Bayesian hierarchical models accounting for spatial variation at both
hydrographic river basin and site levels to jointly estimate the temporal trends in these metrics. We found a
great heterogeneity in the temporal trends at the site level
(community biomass: from `r grbmn`% to +`r grbmx`% per decade, species richness:
`r grrhn` to +`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`,
connectance: `r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig.
S1) which was effectively captured by the hierarchical model ( conditional R2
ranging from `r r2cond$log_bm_std` for community biomass to `r
r2cond$log_rich_std` for species richness, Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of
biodiversity and food web structure. Dashed and solid arrows display
respectively negative and positive the standardized slope coefficients, their
widths being proportional to the absolute values of the coefficients.. The
standardized slope coefficients and their associated 95% confidence intervals
obtained by bootstrapping are displayed along the path. R2 displays the marginal
R-squared of the linear models. (b to h) Direct bivariate relationships derived
from the structural equation model. Points display the temporal trends at each
site estimated with a hierarchical Bayesian model, lines display the linear
relationship estimated with the SEM. The distribution of the site temporal
trends are displayed in Fig. S1. All the effects are reported in Table S3.
Average trophic level is weighted by the biomass of the trophic species.
Redundancy is the proportion of redundant links."

```

```{r sem, fig.cap = sem_cap, out.width = "100%"}
tar_load(p_sem_tps_file)
knitr::include_graphics(p_sem_tps_file)
```

@@

These temporal trend estimations at the site level were the input to our SEM,
allowing us to evaluate the direct effect of the temporal trends of species
richness and community biomass on the temporal trends of food web structure, the
direct effect of  species richness temporal trends on the temporal trends of
community biomass, and the indirect effect of species richness temporal trends
on the temporal trends in food web structure through community biomass. We
detected a strong positive effect of species richness temporal trends on the
temporal trends in community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3). This indicates that communities with decreasing, or increasing,
species richness over time experienced a concomitant decrease, or increase, in
community biomass.

We also find a positive effect of the temporal trend in community biomass on the
temporal trends of connectance, average trophic level, and redundancy (resp.
$r_\delta =$ `r rdtps$ct$bm`, `r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig.
\@ref(fig:sem)b-d). Thus, when communities experienced a decrease, or increase,
in community biomass, they also experienced a concomitant decrease, or increase,
in connectance, average trophic level and redundancy. To assess whether top
trophic level species were driving those relationships, we additionally looked
at the temporal trends in the number and the proportion of piscivorous species.
The temporal trends in community biomass were positively associated with both
the temporal trends in the number and proportion of piscivorous species (Fig.
S2a). Then, communities with increasing biomass have an increase in average
trophic level, in the number and proportion of piscivorous fishes (or
reciprocally for decreasing biomass), suggesting that temporal trends in
community biomass are driven by high trophic level fishes.

@@

Spatial analyses, either using empirical data or the bioenergetic food web
model, largely aligned with temporal trends (Fig. \@ref(fig:std)a and S3),
showing consistent positive effects of species richness on community biomass and
of community biomass on average trophic level and redundancy. Species richness
also exhibited a direct negative effect on connectance and a positive effect on
redundancy, mirroring temporal patterns. However, spatial and temporal analyses
diverged in their effects of species richness on average trophic level, with
spatial data showing a negative effect and temporal trends indicating a positive
one. Detailed statistical effects and comparisons are provided in the Appendix.

@@

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S3, R-squared in Table S4, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Figure S3. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S4).")
```

@@

We selected the sites that have been monitored for at least 10 years in the
1995-2018 period, with the same sampling protocol over time. We selected one
sampling per year by site. In each site, we ensured low temporal heterogeneity
within the sampling period by selecting samplings that were performed in the
same period, leading to a mean standard deviation of sampled month in each site
of three weeks [@danet_species_2021]. Those steps lead to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet)a, Table S1).

@@

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S5). We also computed the marginal ($R^2_m$) and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

Overall, we found that standard deviation attributed to site effects in both
baseline and temporal trends in community structure were much higher than the
ones found for hydrographic basin (Table S5).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were
standardized and the confidence intervals were computed with semEff R package
[@murphy_semeff_2021]. Confidence intervals were obtained by bootstrapping the
linear models 1000 times and computing adjusted bootstrap percentile intervals
(BCa), which are adjusted for the asymmetric distribution of resampling. The marginal
and conditional $R^2$ associated with the linear models of the SEMs were computed
with the Nagelkerke method [@lefcheck_piecewisesem_2016, Table S4].

All the statistical models displayed low multicollinearity (Variance inflation
factors < 3, Table S6). We visually inspected the residuals of the linear
models.

>>>>

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Willem Bonnaffé (d), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (e)"
date: "Research report"
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

(d) Department of Biology, University of Oxford, Oxford, United Kingdom

(e) Sorbonne Université, CNRS, IRD, INRAE, Université Paris Est Créteil, 15
Institute of Ecology and Environmental Sciences of Paris (iEES-Paris), Paris,
France

@@

- Keywords: food web, biodiversity changes, redundancy, ecosystem functioning,
  vulnerability

# Abstract {-}

Biodiversity temporal changes are expected to profoundly impact ecosystem
functioning and stability, yet large-scale empirical evidence remains limited.
Bridging this gap requires aligning biodiversity temporal trends with food web
theory and Biodiversity-Ecosystem Functioning research, which provide strong
predictions about the ecological consequences of biodiversity loss. Most
insights rely on spatial comparisons of ecological communities, assuming they
reflect temporal trends—a space-for-time substitution. Here, we analyze
biodiversity and food web time series from over 400 riverine fish communities
across France (1995–2018) to evaluate how biodiversity temporal trends reshape
food web structure and community functioning. We examine relationships among
species richness, community biomass, and three key food web metrics:
connectance, weighted average trophic level, and trophic pathway redundancy. We
also test the space-for-time hypothesis using spatial gradients and a
theoretical food web model. Structural equation models and robustness analyses
reveal that declines in species richness strongly correlate with reduced
community biomass, with top trophic-level species playing a disproportionate
role in food web changes. Declining biomass is associated with the loss of top
trophic levels and decreased connectance, while temporal changes of species
richness negatively correlates with those of connectance. Both biomass and
species richness declines reduce trophic pathway redundancy, suggesting
biodiversity loss undermines community robustness to future perturbations. Our
findings align with food web theory and Biodiversity-Ecosystem Functioning
research, support the validity of space-for-time approaches for basic food web
metrics, and highlight the relevance of food web metrics as indicators of
ecosystem function and fragility.

# Significance {-}

Understanding how biodiversity loss affects ecological communities is crucial
for predicting and mitigating human impacts on ecosystems. This study provides
comprehensive evidence on how biodiversity changes are reshaping food web
structure using an extensive dataset of riverine fish communities. We
demonstrate that temporal changes in species richness and community biomass
consistently correlate with altered trophic levels, connectance, and the
redundancy of feeding relationships, suggesting energetic constraints and
biodiversity-ecosystem functioning mechanisms jointly shape food web assembly
and disassembly. Importantly, we show that declining biodiversity reduces the
redundancy of trophic pathways, potentially making communities more vulnerable
to future perturbations. These results suggest that food web structure metrics
could serve as valuable indicators for monitoring ecosystem health and
evaluating conservation strategies.

# Introduction

Human pressures are driving profound biodiversity changes, including freshwater
and terrestrial biodiversity declines linked to land degradation
[@newbold_global_2015; @cardinale_is_2018; @danet_past_2024] and reductions in
marine fish biomass due to intensified fishing pressures on predatory species
[@christensen_century_2014]. A central goal of Biodiversity-Ecosystem
Functioning and food web research has been to understand the ecological
consequences of these widespread biodiversity losses for ecosystem functioning
and stability [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @schneider_animal_2016; @dunne_network_2002;
@sole_complexity_2001; @allesina_functional_2009; @montoya_press_2009;
@sanders_trophic_2018]. However, research on biodiversity changes remains
largely disconnected from studies examining their ecological impacts, primarily
due to a lack of simultaneous temporal data on biodiversity and food web
dynamics [@danet_species_2021]. As a result, much of our understanding of the
ecological consequences of biodiversity changes over time relies on experimental
and theoretical approaches [@loreau_biodiversity_2001; @loreau_ecological_2022;
@thebault_food-web_2003; @dunne_network_2002; @sole_complexity_2001;
@allesina_functional_2009]. These approaches often rely on the critical
assumption that spatial comparisons of ecological communities can serve as a
proxy for temporal trends—a space-for-time substitution—when direct temporal
data are scarce [@damgaard_a_2019]. To advance our understanding of the
ecological consequences of biodiversity changes, it is essential to investigate
how these changes are reshaping food web structure in natural settings.

Here, we bridge the gap between biodiversity changes and food web dynamics by
leveraging a unique dataset of riverine fish communities sampled repeatedly over
a 23-year period (1995–2018) across 403 locations in France. By reconstructing
yearly food web structures [@danet_species_2021; @bonnaffe_comparison_2021], we
address three key objectives: (i) documenting how temporal biodiversity trends
reshape food web structure, (ii) connecting these changes to established
theories of food webs and Biodiversity-Ecosystem functioning relationships, and
(iii) assessing the validity of the space-for-time substitution assumption in
the context of biodiversity and food web structure changes.

The structure of ecological networks, which describes the number, distribution,
and strength of interactions among species, is critical for understanding
species coexistence, ecosystem functioning, and stability to perturbations
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024]. Both
theory and empirical evidence demonstrate that species richness and food web
structure influence coexistence [@chesson_interaction_2008], ecosystem
productivity [@thebault_food-web_2003; @wang_biodiversity_2018], temporal
stability [@danet_species_2021; @danet_response_2024;
@eschenbrenner_diversity_2023], and robustness to disturbances
[@sole_complexity_2001; @dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @montoya_press_2009; @sanders_trophic_2018]. For instance,
keystone predators can regulate dominant competitors, preventing
competitive exclusion and maintaining species richness [@heithaus_predicting_2008;
@johnson_rarity_2007; @estes_trophic_2011]. Similarly, food
webs with higher trophic levels are associated with greater ecosystem
productivity [@wang_biodiversity_2018; @schneider_animal_2016;
@wang_intraguild_2019; @nie_will_2023] and community stability
[@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024],
while the number and distribution of trophic links determine resilience to
perturbations [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009;
@allesina_functional_2009; @sanders_trophic_2018]. Understanding how biodiversity changes
reshape ecological networks is therefore essential for quantifying and
mitigating anthropogenic impacts on ecosystems.

Three key components link biodiversity changes to food web structure. First, the
energetic paradigm of food web dynamics posits that the presence of top
consumers depends on sufficient resource availability at lower trophic levels,
driven by bottom-up productivity [@lindeman_trophic-dynamic_1942;
@gravel_trophic_2011]. Thus, the biomass of generalist top consumers is likely
constrained by overall community biomass. Second, connectance—the proportion of
realized trophic links—typically decreases with increasing species richness, as
consumers are limited in the number of resources they can exploit
[@oconnor_general_2017; @dunne_network_2006]. However, higher species richness
is expected to increase the redundancy of trophic pathways, enhancing ecosystem
robustness to species extinctions [@dunne_network_2002; @allesina_who_2004;
@bodini_using_2009; @allesina_functional_2009]. This redundancy underpins the
insurance effect of biodiversity, where greater diversity buffers ecosystems
against perturbations and environmental changes [@yachi_biodiversity_1999;
@loreau_ecological_2022; @danet_response_2024]. Finally, biodiversity-ecosystem
functioning theory suggests that species richness promotes community biomass
through niche complementarity [@loreau_biodiversity_2001]. In food webs, the
presence of higher trophic levels reduces niche overlap among producers, further
boosting productivity and biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023]. Together, these
mechanisms highlight how biodiversity changes can alter trophic link
distribution, robustness, and ecosystem functioning.

To investigate how temporal biodiversity changes correlate with changes in food
webs, we analyzed temporal trends in species richness, biomass, and food web
structure across `r nb_op` sampling events from `r nb_st` fish communities (Fig.
\@ref(fig:matmet)a-c, Table S2). Using structural equation modeling (SEM), we
inferred the relative effects of biomass and species richness temporal trends on
trends of food web metrics, as well as the effect of species richness trends on
biomass trends (Fig. \@ref(fig:matmet)d). Following a classical approach of
network analysis [@dunne_network_2002], we also analyzed the robustness to
species extinction of the meta-food web (comprising all trophic species and
their interactions) to identify species extinction scenarios that could explain
observed variations in food web structure. To test the space-for-time
hypothesis, we compared temporal findings with spatial gradients using two
approaches: (i) replicating our temporal analysis using median biodiversity and
food web metrics across sites to remove temporal trends (Fig. \@ref(fig:matmet)c
& e), and (ii) simulating food web dynamics across gradients of species richness
and primary productivity using a bioenergetic model [@brose_allometric_2006;
@lajaaiti_ecologicalnetworksdynamicsjl_nodate]. These simulations allowed us to
evaluate whether theoretical predictions align with our empirical data, which
are limited to fish biodiversity.

(ref:fig-met-cap) Methodological Approach: (a) We analyzed `r nb_st` stream fish
communities in France, each monitored for at least 10 years. (b) For each site
and sampling event, we reconstructed food webs using a metaweb based on
ontogenetic diet information and predator-prey body size ratios. (c) We computed
temporal trends and median values for community biomass, species richness,
connectance, average trophic level, and redundancy at each site. (d) We then
assessed relationships among these food web descriptors over time and (e) across
spatial gradients.

```{r matmet, fig.cap="(fig-met-cap)", out.width="100%"}
knitr::include_graphics(get_mypath("figures", "mat_met.png"))
```

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species richness,
biomass, connectance, weighted average trophic level, and proportion of
redundant trophic links (hereafter “redundancy”, see Appendix 1,
Fig. S1, Table S2) from the `r nb_st`
sampling sites with 10+ years of community dynamics (Fig. \@ref(fig:matmet)a-c). We used
ontogenic food diet data at the species level and individual fish body size to
reconstruct trophic species food webs, i.e. at the size class species level (See
Methods). We implemented Bayesian hierarchical models accounting for spatial
variation at both hydrographic river basin and site levels to jointly estimate
the temporal trends in these metrics (see Methods). We found a great
heterogeneity in the temporal trends at the site level (community biomass: from
`r grbmn`% to +`r grbmx`% per decade, species richness: `r grrhn` to
+`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig.
S2) which was effectively captured by the hierarchical model (conditional R2
ranging from `r r2cond$log_bm_std` for community biomass to
`r r2cond$log_rich_std` for species richness, Table S3).

(ref:sem-cap) (a) Structural equation models linking the temporal trends of
biodiversity and food web structure. Dashed and solid arrows display
respectively negative and positive the standardized slope coefficients, their
widths being proportional to the absolute values of the coefficients. The
standardized slope coefficients and their associated 95% confidence intervals
obtained by bootstrapping are displayed along the path. $R^2$ displays the marginal
R-squared of the linear models. (b to h) Direct bivariate relationships derived
from the structural equation model. Points display the temporal trends at each
site estimated with a hierarchical Bayesian model, lines display the linear
relationship estimated with the SEM. The distribution of the site temporal
trends are displayed in Fig. S2. All the effects are reported in Table S4.
Average trophic level is weighted by the biomass of the trophic species.
Redundancy is the proportion of redundant links.

```{r sem, fig.cap = "(ref:sem-cap)", out.width = "100%"}
tar_load(p_sem_tps_file)
knitr::include_graphics(p_sem_tps_file)
```

@@

These temporal trend estimations at the site level were the input to our SEM
which accounts for uncertainties in trend estimation (see Methods). It allows us
to evaluate the direct effect of the temporal trends of species richness and
community biomass on the temporal trends of food web structure, the direct
effect of species richness temporal trends on the temporal trends of community
biomass, and the indirect effect of species richness temporal trends on the
temporal trends in food web structure through community biomass. We detected a
strong positive effect of the temporal trends in species richness on the
temporal trends in community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S4). This indicates that communities with decreasing, or increasing,
species richness over time experienced a concomitant decrease, or increase, in
community biomass.

We also find positive effects of the temporal trend in community biomass on the
temporal trends of connectance, average trophic level, and redundancy (resp.
$r_\delta =$ `r rdtps$ct$bm`, `r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig.
\@ref(fig:sem)b-d). Thus, when communities experienced a decrease, or increase,
in community biomass, they also experienced a concomitant decrease, or increase,
in connectance, average trophic level and redundancy. To assess whether top
trophic level species were linked to those relationships, we additionally looked
at the temporal trends in the number and the proportion of piscivorous species.
The temporal trends in community biomass were positively associated with the
temporal trends in both the number and the proportion of piscivorous species
(Fig. S3a). It suggests that high trophic level fishes are strongly linked to
temporal trends in community biomass.

HERE

@@

Spatial analyses, either using empirical data or the bioenergetic food web
model, largely aligned with temporal trends (Fig. \@ref(fig:std)a and S4),
showing consistent positive effects of species richness on community biomass and
of community biomass on average trophic level and redundancy. Species richness
also exhibited a direct negative effect on connectance and a positive effect on
redundancy, mirroring temporal patterns. However, spatial and temporal analyses
diverged in their effects of species richness on average trophic level, with
spatial data showing a negative effect and temporal trends indicating a positive
one. Detailed statistical effects and comparisons are provided in the Appendix.

@@

```{r}
std_cap <- paste0( "a, Direct and total effects of community biomass and species
  richness trends on food web structure trends. Standardized coefficients (r)
  derived from the temporal, spatial, and bioenergetic model Structural Equation
  Models. Points and lines display the estimated effects and their confidence
  intervals at 95\\%. Values are displayed in Table S4, R-squared in Table S5, and
  the visualization of the bivariate relationships for the spatial and
  bioenergetic model SEMs are displayed in Fig. S4. Average trophic level is
  weighted by the biomass of the trophic species. Redundancy is the proportion
  of redundant links. b, Effect of species extinctions on the structure of the
  empirical metaweb according to three scenarios. In the Decreasing and
  Increasing scenarios, species were ordered by their trophic level and removed
  from the metaweb starting respectively from the highest and lowest trophic
  level. Species were removed in a random sequence (50 sequences, grey dots) for
  the random scenario, the black line displaying the average predicted by a GAM
  linear model. For the extinction sequence, the average trophic level was not
  weighted by the biomass of species as this information is not available at the
  metaweb scale. Similar results were obtained with trophic species (i.e. node)
  extinction (Fig. S5).")
```

@@

We selected the sites that have been monitored for at least 10 years in the
1995-2018 period, with the same sampling protocol over time. We selected one
sampling per year by site. In each site, we ensured low temporal heterogeneity
within the sampling period by selecting samplings that were performed in the
same period, leading to a mean standard deviation of sampled month in each site
of three weeks [@danet_species_2021]. Those steps lead to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet)a, Table S2).

@@

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S6). We also computed the marginal ($R^2_m$) and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

Overall, we found that standard deviation attributed to site effects in both
baseline and temporal trends in community structure were much higher than the
ones found for hydrographic basin (Table S6).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were
standardized and the confidence intervals were computed with semEff R package
[@murphy_semeff_2021]. Confidence intervals were obtained by bootstrapping the
linear models 1000 times and computing adjusted bootstrap percentile intervals
(BCa), which are adjusted for the asymmetric distribution of resampling. The marginal
and conditional $R^2$ associated with the linear models of the SEMs were computed
with the Nagelkerke method [@lefcheck_piecewisesem_2016, Table S5].

All the statistical models displayed low multicollinearity (Variance inflation
factors < 3, Table S7). We visually inspected the residuals of the linear
models.
